UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-04-10 ms/main.tex
update MS with new figures and text

\begin{abstract}
{\large Abstract.}

Genetic structure in large cohorts results from technical, sampling, and demographic variation. Identifying structure through visualization is typically the first step in most genomic analyses. Existing exploration methods struggle with unbalanced sampling and the many scales of population structure.

We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly illustrate population structure in large cohorts and capture their relationships on local and global scales. Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP effectively clusters closely related individuals while placing them in a global continuum of genetic variation. This approach reveals previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population. 
This opens new lines of investigation for demographic research and statistical genetics. Given its small computational cost, PCA-UMAP also provides a general-purpose approach to exploratory analysis in population-scale datasets.
\end{abstract}

\keywords{genomics $|$ genetics $|$ population structure $|$ ethnicity $|$ ancestry $|$ machine learning $|$ data visualization $|$ dimension reduction}

@@

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

\subsection*{Author summary}
%We ask that all authors of research articles include a 150?200 word non-technical summary of the work as part of the manuscript to immediately follow the abstract. This text is subject to editorial change, should be written in the first-person voice, and should be distinct from the scientific abstract.

%Aim to highlight where your work fits within a broader context; present the significance or possible implications of your work simply and objectively; and avoid the use of acronyms and complex terminology wherever possible. The goal is to make your findings accessible to a wide audience that includes both scientists and non-scientists.

%Authors may benefit from consulting with a science writer or press officer to ensure they effectively communicate their findings to a general audience.
Because of geographic isolation, individuals tend to be more genetically related to people living nearby than to people living far. This is an example of population structure, a situation where a large population contains subgroups that share more than the average amount of DNA. This structure can tell us about human history, and it can also have a large effect on medical studies.  We use a newly developed method (UMAP) to visualize population structure from three genomic datasets. Using genotype data alone, we reveal numerous subgroups related to ancestry and correlated with traits such as white blood cell count, height, and FEV1, a measure used to detect airway obstruction. We demonstrate that UMAP reveals previously unobserved patterns and fine-scale structure. We show that visualizations work especially well in large datasets containing populations with diverse backgrounds, which are rapidly becoming more common, and that unlike other visualization methods, we can preserve intuitive connections between populations that reflect their shared ancestries. The combination of these results and the effectiveness of the strategy on large and diverse datasets make this an important approach for exploratory analysis for geneticists studying ancestral events and phenotype distributions.

\subsection*{Introduction}

Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference}.

Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations. If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of derived alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ typically larger than $100,000$. Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.

Principal component analysis (PCA) is often the first dimensional reduction tool used for genomic data. It identifies and ranks directions in genotype space that explain most-to-least variance among individuals. Positions of individuals along directions of highest variance can then be used to summarize individual genotypes. PCA coordinates have natural genealogical interpretations in terms of times to a most recent common ancestor (TMRCA) \cite{mcvean2009genealogical}, and are used empirically to reveal admixture \cite{brisbin2012pcadmix}, continuous isolation-by-distance \cite{novembre2008europe, nelson2008population}, as well as technical artefacts. PCA coordinates are particularly well-suited to correct for population structure in GWAS~\cite{eigen2006}.

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=0.7\textwidth]{images/montage_1000G_labels_axes.pdf}
    \caption{Four methods of dimension reduction of 1KGP genotype data with population labels (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian/Native American, and European continental ancestry. Discarding lower-variance PCs leads to overlap of populations with no close affinity, such as Central and South American populations with South Asians. (ii) t-SNE forms groups corresponding to continents, with some overlap between European and Central and South American people. Smaller subgroups are visible within continental clusters. The cloud of peripheral points results from the method's poor convergence. (iii) UMAP forms distinct clusters related to continent with clearly defined subgroups. Japanese, Finnish, Luhya, and some Punjabi and Telugu populations form separate clusters consistent with their population history\cite{10002015global}. (iv) UMAP on the first 15 principal components forms fine-scale clusters for individual populations. Groups closely related by ancestry or geography, such as African Caribbean/African American, Spanish/Italian, and Kinh/Dai populations cluster together. Results using t-SNE on principal components are presented in figure~\ref{fig:supp_megamontage_pc2_9}. Axes in UMAP and t-SNE are arbitrary.
    ACB, African Caribbean in Barbados;
    ASW, African Ancestry in Southwest US;
    BEB, Bengali;
    CDX, Chinese Dai;
    CEU, Utah residents with Northern/Western European ancestry;
    CHB, Han Chinese;
    CHS, Southern Han Chinese;
    CLM, Colombian in Medellin, Colombia;
    ESN, Esan in Nigeria;
    FIN, Finnish;
    GBR, British in England and Scotland;
    GWD, Gambian;
    GTH, Gujarati;
    IBS, Iberian in Spain;
    ITU, Indian Telugu in the UK;
    JPT, Japanese;
    KHV, Kinh in Vietnam;
    LWK, Luhya in Kenya;
    MSL, Mende in Sierra Leone;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PJL, Punjabi in Lahore, Pakistan;
    PUR, Puerto Rican;
    STU, Sri Lankan Tamil in the UK;
    TSI, Tuscani in Italy;
    YRI, Yoruba in Nigeria}
    \label{fig:1000g_compare}
\end{figure*}

As sample sizes increase, the amount of information encoded in the lower-variance principal components increases, and researchers typically examine multiple two-dimensional projections to get a sense of the data. While many features of the data can be identified in this manner, other features may be hidden by the projections or hard to interpret.

To display more of the high-dimensional features of the data in a two dimensional figure,  we can use non-linear transformations that seek to preserve the local structure of the data. A popular method for visualization is t-distributed stochastic neighbour embedding (t-SNE)\cite{maaten2008visualizing}. Whereas PC projection is designed to capture as much variance as possible for a linear transformation, t-SNE seeks a low-dimensional representations of the data that preserves pairwise distances, using a probabilistic heuristic to weight mismatches. t-SNE has been used before to visualize SNPs\cite{platzer2013visualization}. Using data from the 1000 Genomes Project (1KGP)\cite{10002015global}, it groups individuals corresponding roughly to their continent of origin, with smaller ethnic sub-groups visible within the larger continental clusters\cite{li2017tsne}. However, t-SNE struggles with very large datasets, when a large number of locally optimal configurations make convergence to a globally satisfying solution difficult.

Uniform Manifold Approximation and Projection (UMAP) is a new dimension reduction technique grounded in Riemannian geometry, algebraic topology, and category theory, and designed to model and preserve the high-dimensional topology of data points in the low-dimensional space\cite{2018arXivUMAP}. The assumption behind UMAP is that data are uniformly distributed on local manifolds in high-dimensional space, which can be approximated as fuzzy sets that are patched together to form a topological representation. One can then construct a low-dimensional topological representation that minimizes the differences between the two representations. With genotype data, UMAP creates a neighbourhood around each individual's genetic coordinates and identifies a pre-selected number of neighbours to build high-dimensional manifolds. The end result is a low-dimension representation that groups genetically similar individuals together on a local scale while preserving long-range topological connections to more distantly related individuals. The method has been successfully applied to single-cell RNA sequencing datasets\cite{umap2018singlecell}.

A common practice in dimensional reduction is to first apply PCA to reduce the number of dimensions before performing nonlinear dimensional reduction. In addition to being computationally advantageous, this discards statistical noise that can confound nonlinear approaches: Population structure arising from $n$ isolated randomly-mating demes can be described by the leading $n-1$ PCs, with the following PCs describing stochastic variation in relatedness~\cite{eigen2006}. Selecting the leading PCs therefore has potential to extract meaningful population structure while filtering out stochastic noise.

We explore different strategies to pre-process the data and investigate discrete and continuous population structure patterns present in large datasets of human genotypes: the 1KGP, the Health and Retirement Study (HRS)\cite{juster1995overview}, and the UK BioBank (UKBB)\cite{sudlow2015uk}.

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=0.6\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
    \end{subfigure}
    \caption{Applying UMAP to the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted). Coloring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster. Six 1KGP populations are presented:
    CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS colored by estimated admixture proportions.}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
\end{figure*}

\subsection*{Results}
\subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}. Figure~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and UMAP with PCA pre-processing. Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters. Using either on the top principal components leads to more distinct population clusters and less defined continental structure (see figure \ref{fig:supp_megamontage_pc10_50} for tSNE). Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; further components gradually approach results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}).

Focusing on UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), we also find several population clusters that reflect shared ancestries. British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry. Toscani and Iberian individuals form a group reflecting their Mediterranean heritage. African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster. The East Asian super-population forms three sub-populations split by geography: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population. Looser geographical groupings include Colombians and Peruvians, and the Esan and Yoruba populations of Nigeria; both groupings appear as connected sub-clusters. The South Asian super-population also forms a loose grouping.

Only a few individuals cluster differently than the majority of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority. One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone. Only two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering is robust to, e.g., the choice of the number of PCs considered (see figure ~\ref{fig:supp_megamontage_pc10_50}). Analysis by 23andMe determined the distinct Gujarati cluster mostly have the last name Patel and suggested their grouping is the result of endogamy~\cite{23andme}.

Finally, contrasting UMAP and t-SNE, we find that UMAP preserves more of the global structure of the data than t-SNE, and is more robust to choices of data pre-processing (figure~\ref{fig:supp_megamontage_pc10_50}).

\subsubsection*{The genetic continuum of admixed populations} The 1KGP sampled individuals from relatively distinct population groups across the world, which makes the data particularly easy to cluster. Most medical cohorts comprise larger numbers of individuals sampled across extended geographical areas.

For example, the HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}. We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g. White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau. Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian/Native American, and European populations using RFMIX \cite{Maples:2013fia}.
We have scaled these three proportions to values between 0 and 255, to color individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry. Using the first 10 principal components and UMAP, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.

@@

To investigate possible ancestries related to populations in the 1KGP we took two approaches. In the first we generated PC axes and a UMAP embedding using UMAP and 1KGP data together (figure~\ref{fig:supp_hrs_1000g}). In the second we used the PC axes and UMAP embedding generated in figure~\ref{fig:umap_hrs_admix} and projected 1KGP data onto it (figure~\ref{fig:supp_hrs_1kgp_projected}). Both approaches reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population. One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.

\subsubsection*{Regional patterns in the American Hispanic population} In contrast to the WNH individuals, applying UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region. One separate cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States. This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in figure~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6. Even though continental admixture patterns do correlate with UMAP position (figure~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster. Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding. The cluster possibly comprises the Hispano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations. A recent preprint discusses the Mountain Region Hispanics and provides a more detailed historical description \cite{Jordan333609}.

@@

\begin{figure*}%
\centering
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.pdf}%
\caption{Principal components 1 and 2}%
\label{fig:pc_ukbb}%
\end{subfigure}\hfill%
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.pdf}%
\caption{UMAP on first 10 principal components}%
\label{fig:umap_ukbb_pc10}%
\end{subfigure}%
\caption{The UKBB projected onto two dimensions, colored by self-reported ethnic background. (a) The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations. (b) UMAP on the first 10 principal components. The cluster of White British and White Irish individuals is greatly expanded, with the Irish forming a distinct sub cluster mixed with the White British population. South Asian and East Asian individuals form their separate clusters, as do individuals of African or Caribbean backgrounds. Population clusters are connected by "trails" comprised of large proportions of individuals with mixed backgrounds.
BA, Black African; 
BC, Black Caribbean; 
BG, Bangladeshi; 
CHN, Chinese; 
IND, Indian;
PK, Pakistani;
WB, White British;
WI, White Irish;
WBC, White and Black Caribbean; 
WBA, White and Black African; 
WAA, White and Asian;
AAB, Any other Asian Background; 
ABB, Any other Black Background;
AWB, Any other White Background;
AMB, Any other Mixed Background;
OEG, Other ethnic group. 
}
\label{fig:fig_ukbb}
\end{figure*}

The largest body in the figure consists of the White British and Irish populations. The Irish population concentrates in a portion of this group, but many individuals are also scattered throughout the British-identified population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within appears to be a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well. The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals. These trails come together in a nexus of individuals with a variety of ethnicities; many claim mixed ancestry, and there are clear groups of individuals who belong to an "other ethnic group". Although their ethnicities are unknown to us, given their proximity to African, South Asian, and White individuals, possible candidates for these groups are North African, Middle Eastern, and West Asian backgrounds. Additionally, there are many individuals whose ethnicity is White but neither British nor Irish (AWB) forming clusters distinct from the British and Irish cluster.

>>>>

%An Analysis is a new analysis of existing data (typically large genomic, transcriptomic or proteomic data sets from arrays or other high-throughput platforms) or new data obtained in a comparative analysis of technologies that lead to novel and arresting conclusions of importance to a broad audience. The main text (excluding abstract, Methods, references and figure legends) is approximately 3,000 words. The abstract is typically 100-150 words, unreferenced. Analyses have no more than 6 display items (figures and/or tables). An introduction (without heading) is followed by sections headed Results, Discussion and online Methods. The Results and online Methods should be divided by topical subheadings; the Discussion does not contain subheadings. As a guideline, Analysis allow up to 50 references.

\begin{abstract}
{\large Abstract.}
%The abstract is typically 100-150 words, unreferenced. 
Improvements in sequencing have generated massive and diverse genomic datasets of populations. Dimension reduction is necessary to condense and analyze the data. Uniform manifold approximation and projection (UMAP) is a non-linear dimension reduction tool recently made available. Here we apply UMAP to three well-studied genotype datasets and discover overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography, genotypes, and phenotypes in the UK population, and cryptic structure in the Thousand Genomes Project data. We demonstrate this approach outperforms other standard methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry population-scale datasets.
\end{abstract}

@@

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference}.

Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations. If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ typically larger than $100,000$. Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.

Principal component analysis (PCA) is often the first dimensional reduction tool used for genomic data. It identifies and ranks directions in genotype space that explain most-to-least variance among individuals. Positions of individuals along directions of highest variance can then be used to summarize individual genotypes. PCA coordinates have natural genealogical interpretations in terms of expected times to a most recent common ancestor (TMRCA) \cite{mcvean2009genealogical}, and are used empirically to reveal admixture \cite{brisbin2012pcadmix}, continuous isolation-by-distance \cite{novembre2008europe, nelson2008population}, as well as technical artefacts. PCA coordinates are particularly well-suited to correct for population structure in GWAS~\cite{eigen2006}.

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=0.7\textwidth]{images/montage_1000G_labels_axes.pdf}
    \caption{Four methods of dimension reduction of 1KGP genotype data with population labels (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian/Native American, and European continental ancestry. Discarding lower-variance PCs leads to overlap of populations with no close affinity, such as Central and South American populations with South Asians. (ii) t-SNE forms groups corresponding to continents, with some overlap between European and Central and South American people. Smaller subgroups are visible within continental clusters. The cloud of peripheral points results from the method's poor convergence. (iii) UMAP forms distinct clusters related to continent with clearly defined subgroups. Japanese, Finnish, Luhya, and some Punjabi and Telugu populations form separate clusters consistent with their population history\cite{10002015global}. (iv) UMAP on the first 15 principal components forms fine-scale clusters for individual populations. Groups closely related by ancestry or geography, such as African Caribbean/African American, Spanish/Italian, and Kinh/Dai populations cluster together. Results using t-SNE on principal components are presented in figure~\ref{fig:supp_megamontage_pc2_9}. Axes in UMAP and t-SNE are arbitrary. Since the algorithms prioritize local distances, long distances between clusters are not meaningful.
    ACB, African Caribbean in Barbados;
    ASW, African Ancestry in Southwest US;
    BEB, Bengali;
    CDX, Chinese Dai;
    CEU, Utah residents with Northern/Western European ancestry;
    CHB, Han Chinese;
    CHS, Southern Han Chinese;
    CLM, Colombian in Medellin, Colombia;
    ESN, Esan in Nigeria;
    FIN, Finnish;
    GBR, British in England and Scotland;
    GWD, Gambian;
    GTH, Gujarati;
    IBS, Iberian in Spain;
    ITU, Indian Telugu in the UK;
    JPT, Japanese;
    KHV, Kinh in Vietnam;
    LWK, Luhya in Kenya;
    MSL, Mende in Sierra Leone;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PJL, Punjabi in Lahore, Pakistan;
    PUR, Puerto Rican;
    STU, Sri Lankan Tamil in the UK;
    TSI, Tuscani in Italy;
    YRI, Yoruba in Nigeria}
    \label{fig:1000g_compare}
\end{figure*}

The amount of information encoded in the lower-variance principal components increases with sample size, so researchers typically examine multiple two-dimensional projections to explore data. Some features may be hidden by the projections or hard to interpret. To display more of the high-dimensional features of the data in a two dimensional figure, we can use non-linear transformations that seek to preserve the local structure of the data. A popular method for visualization is t-distributed stochastic neighbour embedding (t-SNE)\cite{maaten2008visualizing}. t-SNE has been used before to visualize SNPs\cite{platzer2013visualization}; using data from the 1000 Genomes Project (1KGP)\cite{10002015global}, it groups individuals corresponding roughly to their continent of origin, with smaller ethnic sub-groups visible within the larger continental clusters\cite{li2017tsne}. However, t-SNE struggles with very large datasets, when a large number of locally optimal configurations make convergence to a globally satisfying solution difficult.

Uniform Manifold Approximation and Projection (UMAP) is a dimension reduction technique designed to model and preserve the high-dimensional topology of data points in the low-dimensional space\cite{2018arXivUMAP}. With genotype data, UMAP creates a neighbourhood around each individual's genetic coordinates and identifies a pre-selected number of neighbours to build high-dimensional manifolds. The end result is a low-dimension representation that groups genetically similar individuals together on a local scale while preserving long-range topological connections to more distantly related individuals. The method has been successfully applied to single-cell RNA sequencing datasets\cite{umap2018singlecell}.

A common practice in dimensional reduction is to first apply PCA. In addition to being computationally advantageous, this discards noise that can confound nonlinear approaches: population structure arising from $n$ isolated randomly-mating demes can be described by the leading $n-1$ PCs, with the following PCs describing stochastic variation in relatedness~\cite{eigen2006}. Selecting the leading PCs therefore has potential to extract meaningful population structure while filtering out stochastic noise. We explore different strategies to pre-process the data and investigate discrete and continuous population structure patterns present in large datasets of human genotypes: the 1KGP, the Health and Retirement Study (HRS)\cite{juster1995overview}, and the UK BioBank (UKBB)\cite{sudlow2015uk}, and compare UMAP's performance to t-SNE.

\begin{figure*}
    \centering
    \begin{subfigure}{0.5\columnwidth}
    \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
    \caption{Subset of Hispanic population of the HRS}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
    \end{subfigure}\hfill
        \begin{subfigure}{0.5\columnwidth}
    \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC8_NC2_NN15_MD05_20193117303txt_eth_for_ms_lw2.jpeg}
    \caption{Subset of the Asian population of the UKBB}
        \label{fig:umap_ukbb_indian}
    \end{subfigure}
    \caption{Applying UMAP to subets of data can reveal deep population structure. (a) UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted). Colouring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster. Six 1KGP populations are presented:
    CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS colored by estimated admixture proportions. (b) UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters. Almost all individuals in the highlighted cluster are Indian individuals born in Kenya.}
    \label{fig:subpops}
\end{figure*}

\subsection*{Results}
\subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}. Figure~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and UMAP with PCA pre-processing. Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters. Using either on the top principal components leads to more distinct population clusters and less defined continental structure. Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}). Patterns are also evident when using UMAP on the last components (\ref{fig:supp_1kgp_3350}).

Focusing on UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), we find several population clusters that reflect shared ancestries. British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry. Toscani and Iberian individuals form a group reflecting their Mediterranean heritage. African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster. Three East Asian groups appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population. Looser geographical groupings include Colombians and Peruvians, and the Esan and Yoruba populations of Nigeria; both groupings appear as connected sub-clusters. The South Asian populations also form a loose grouping.

Only a few individuals cluster differently than the majority of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority. One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}). Analysis of the cluster by 23andMe suggested the Gujarati grouping is the result of endogamy~\cite{23andme}.

\subsubsection*{The genetic continuum of admixed populations} The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form iclusters. Most medical cohorts comprise larger numbers of individuals sampled across extended geographical areas. For example, the HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}. We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g. White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau. Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian/Native American, and European populations using RFMIX \cite{Maples:2013fia}.
We have scaled these three proportions to values between 0 and 255, to color individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry. Using UMAP on the first 10 principal components, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.

@@

To investigate possible ancestries related to populations in the 1KGP we took two approaches. In the first we generated PC axes and a UMAP embedding using HRS and 1KGP data together (figure~\ref{fig:supp_hrs_1000g}). In the second we used the PC axes and UMAP embedding generated in figure~\ref{fig:umap_hrs_admix} and projected 1KGP data onto it (figure~\ref{fig:supp_hrs_1kgp_projected}). Both approaches reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population. One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.

\subsubsection*{Deep population structure in the American Hispanic subpopulation} In contrast to the WNH individuals, applying UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region. One separate cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States. This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in figure~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6. Even though continental admixture patterns do correlate with UMAP position (figure~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster. Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding. The cluster possibly comprises the Hispano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations. A recent preprint discusses the Mountain Region Hispanics and provides a more detailed historical description \cite{Jordan333609}.

@@

\begin{figure*}%
\centering
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.jpeg}%
\caption{Principal components 1 and 2}%
\label{fig:pc_ukbb}%
\end{subfigure}\hfill%
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.pdf}%
\caption{UMAP on first 10 principal components}%
\label{fig:umap_ukbb_pc10}%
\end{subfigure}%
\caption{The UKBB projected onto two dimensions, colored by self-reported ethnic background. (a) The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations. (b) UMAP on the first 10 principal components. The cluster of White British and White Irish individuals is greatly expanded, with the Irish forming a distinct sub cluster mixed with the White British population. South Asian and East Asian individuals form their separate clusters, as do individuals of African or Caribbean backgrounds. Population clusters are connected by "trails" comprised of large proportions of individuals with mixed backgrounds.
BA, Black African; 
BC, Black Caribbean; 
BG, Bangladeshi; 
CHN, Chinese; 
IND, Indian;
PK, Pakistani;
WB, White British;
WI, White Irish;
WBC, White and Black Caribbean; 
WBA, White and Black African; 
WAA, White and Asian;
AAB, Any other Asian Background; 
ABB, Any other Black Background;
AWB, Any other White Background;
AMB, Any other Mixed Background;
OEG, Other ethnic group. 
}
\label{fig:fig_ukbb}
\end{figure*}

The largest body in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations. The Irish population concentrates in one part of this group, but many individuals are also scattered throughout the British-identified population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well. The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals. These trails come together in a nexus of individuals with a variety of ethnicities. Many claim mixed ancestry, and there are clear groups of individuals who belong to an "other ethnic group"; some of these were identified using country of birth in figure~\ref{fig:supp_ukbb_cob} and appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.
